---
title: "03_Meta-Analysis"
format: html
---

# Load libraries
```{r}
library(vroom)
library(levinmisc)
library(purrr)
library(dplyr)
library(tidyr)

remotes::install_github("lcpilling/gwasRtools")
library(gwasRtools)
```

# QC 

Example of a QC check - calculate gc lambda

```{r}
calculate_lambda_gc <- function(df, PVALUE, EFF_ALL_FREQ, MAF_FILTER = FALSE, MAC_FILTER = FALSE) {
  # Ensure that the 'PVALUE' column exists in the data frame
  if (!"PVALUE" %in% names(df)) {
    stop("The data frame must contain a 'PVALUE' column.")
  }
  
  # Check for 'EFF_ALL_FREQ' column when MAF_FILTER is TRUE
  if (MAF_FILTER && !"EFF_ALL_FREQ" %in% names(df)) {
    stop("The data frame must contain an 'EFF_ALL_FREQ' column for MAF filtering.")
  }
  
  # Apply MAF filter if MAF_FILTER is TRUE
  if (MAF_FILTER) {
    df <- df[df$EFF_ALL_FREQ > 0.01, ]
  }
  
  # Check for 'CASE_MAC' column when MAC_FILTER is TRUE
  if (MAC_FILTER && !"CASE_MAC" %in% names(df)) {
    stop("The data frame must contain a 'CASE_MAC' column for Case MAC filtering.")
  }
  
  # Apply CASE_MAC filter if MAC_FILTER is TRUE
  if (MAC_FILTER) {
    df <- df[df$CASE_MAC > 25, ]
  }
  
  # Calculate the chi-squared values
  chisq_FG_EUR <- qchisq(1 - df$PVALUE, 1)
  
  # Calculate lambda gc (λgc)
  lambda_gc <- median(chisq_FG_EUR) / qchisq(0.5, 1)
  
  return(lambda_gc)
}
```




# MR-MEGA

 Make new files
This is an example of how to format binary outcome data in a way that works for both ML's wrapper, and can be directly used for MR-MEGA for binary outcome applications.
Consider what this function does, and what you actually need to accomplish your goals.

Format GWAS Inspector Output for MR-MEGA
```{r}

inspect_to_mega <- function(infile, outfile) {
  vroom::vroom(infile) %>%
    mutate(
    OR = exp(EFFECT),  # Calculate the Odds Ratio
    OR_95L = exp(EFFECT - 1.96 * STDERR),  # Lower 95% CI
    OR_95U = exp(EFFECT + 1.96 * STDERR), # Upper 95% CI
    allele_count = (AF * N_TOTAL),
    N_EFF = (4 / (1/N_CASE + 1/N_CONTROL))) %>% # Add effective sample size
  filter(allele_count >=50) %>% 
  select (MARKERNAME = varID, EA = EFFECT_ALL, NEA = OTHER_ALL, OR, OR_95L, OR_95U,  EAF = EFF_ALL_FREQ, N = N_EFF, CHROMOSOME = CHR, POSITION, EFFECT, STDERR, PVALUE) %>% 
  drop_na() %>%
  write_tsv(.,outfile)
}

```


```{r}
mr_mega_list <- list.files("<path to formatted summary statistics>", full.names=TRUE)
mr_mega_bin = "/project/voltron//Applications/MR-MEGA/MR-MEGA_v0.2/MR-MEGA" # Specific to LPC. For those using LPC, use this exact file path. 
```
```{r}
#Example
output_mr_mega <-  mr_mega(
  mr_mega_list,
  mr_mega_bin,
  marker_col = MARKERNAME,
  chr_col = CHROMOSOME,
  pos_col = POSITION,
  effect_allele_col = EA,
  other_allele_col = NEA,
  eaf_col = EAF,
  beta_col = EFFECT,
  se_col = STDERR,
  p_value_col = PVALUE,
  n_pcs = 3) # "Default = 4. Please note that the PC count must be < cohort count - 2. Therefore, if five cohorts have been used in the analyse, then the maximum number of PC-s can be two!"
```

```{r}
process_mrmega<- function(df) {
  df %>%
    mutate(p_value_association = pchisq(chisq_association, ndf_association, lower.tail = FALSE)) %>%
    mutate(p_value_ancestry_het = pchisq(chisq_ancestry_het, ndf_ancestry_het, lower.tail = FALSE)) %>%
    mutate(p_value_residual_het = pchisq(chisq_residual_het, ndf_residual_het, lower.tail = FALSE)) %>% 
    filter(chromosome %in% 1:22) %>% 
    annotate_rsids(chrom_col = chromosome, pos_col = position, dbSNP = SNPlocs.Hsapiens.dbSNP144.GRCh38::SNPlocs.Hsapiens.dbSNP144.GRCh38)
  
}

# Lead variants
extract_lead_loci_mega <- function(df, n_study_filter = 2, distance = 1000000, min_locus_snps = 2) {
  df %>%
    filter(p_value_association < 5e-8) %>%
    filter(ncohort >= n_study_filter) %>%
    mutate(chromosome = as.numeric(chromosome)) %>%
    collect() %>%
    tidygenomics::genome_cluster(by = c("chromosome", "position", "position"), max_distance = distance) %>%
    add_count(cluster_id, name = "cluster_n") %>%
    filter(cluster_n >= min_locus_snps) %>%
    group_by(cluster_id) %>%
    slice_min(p_value_association, with_ties = FALSE) %>%
      mutate(signif_het_bonferroni = case_when(
      p_value_ancestry_het < 0.05/nrow(.) ~ TRUE,
      TRUE ~ FALSE
    )) %>%
    ungroup %>%
    arrange(p_value_association) 
}


```

### Find lead loci from MR-MEGA
```{r}

tophits_MR_MEGA <- output_mr_mega %>% 
  process_mrmega() %>% 
  extract_lead_loci_mega()
```





# METAL

Example of running meta on all files
```{r}
all_meta_files <- list.files(
  "<path to output folder from GWAS inspector",
  pattern = "\\.txt\\.gz$",
  full.names = TRUE
)
```

```{r}
metal_config(
  config_name = "Your_name_here",
  output_dir = "/project/damrauer_shared/Users/sabramow/CAD/METAL/Output_2024",
  study_files = all_meta_files,
  MARKERLABEL = "varID",
  CHROMOSOMELABEL = "CHR",
  POSITIONLABEL = "POSITION",
  EFFECT_ALLELE = "EFFECT_ALL",
  OTHER_ALLELE = "OTHER_ALL",
  EFFECTLABEL = "EFFECT",
  STDERR = "STDERR",
  FREQLABEL = "EFF_ALL_FREQ",
  NCASE = "N_CASE",
  NCONTROL = "N_CONTROL",
  SAMPLESIZE = "N_TOTAL"
  
  )


```

```{r}
# Keep metal_path the same
metal_run(config_file = "path which is output by running the above code", metal_path = "/project/voltron/Applications/METAL/metal-2020-05-05")
```


#### Top hits METAL
```{r}

extract_lead_loci_metal <- function(df, distance = 1000000, min_locus_snps = 2) {
  df %>%
    collect() %>%
    tidygenomics::genome_cluster(by = c("Chromosome", "Position", "Position"), max_distance = distance) %>%
    add_count(cluster_id, name = "cluster_n") %>%
    filter(cluster_n >= min_locus_snps) %>% 
    group_by(cluster_id) %>%
    slice_min(P.value, with_ties = FALSE) %>% 
    ungroup()

}
```


```{r}
# This is just an example; to run faster, set ap value filter earlier in the code so you don't need to annotate everything
top_hits_METAL <- df %>% 
  filter(Chromosome %in% 1:22) %>% 
  filter(P.value < 5e-8) %>% 
  mutate_at(vars(Allele1, Allele2), toupper) %>% 
  annotate_rsids(chrom_col = Chromosome, pos_col = Position, dbSNP = SNPlocs.Hsapiens.dbSNP144.GRCh38::SNPlocs.Hsapiens.dbSNP144.GRCh38) %>% 
  extract_lead_loci_metal(.)
  
```


# Manhattan Plotting and qq plots
recommend gg_manhattan_df from levinmisc, or qqman package
don't forget to specify build!
```{r}


```


